---
title: "Workshop_1"
author: "Kelsey Ogren"
date: "2024-09-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#MB5370 Data Science in R Module Workshop 1 - Getting Started

# Install and load tidyverse packages
```{r}
#install.packages("tidyverse") # Delete this line once installed
library("tidyverse")
```


# Load the data (mpg is built into ggplot2)
```{r}
data(mpg)
mpg
```


# Quick data checks
```{r}
head(mpg)
glimpse(mpg)
summary(mpg)
```


# Create first plot
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

The plot shows a negative relationship between engine size (displ) and fuel efficiency (hwy). In other words, cars with big engines use more fuel. What does this say about fuel efficiency and engine size?



# Understanding grammar of graphics
What happens if we just try to run ggplot on its own?
```{r}
ggplot() 
# Or with just the data specified?
ggplot(data = mpg)
```

We need to map some aesthetics!!

# When you’re creating a plot, you essentially need two attributes of a plot: a geom and aesthetics.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```


# Change point colour by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```

# Change point size by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```
Note the warning!!!


# Change transparency (alpha) by class
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```
Another warning!!
Question: When would using alpha (or size be appropriate??)

** Using alpha or size for the data would work well if the data are counts and it is important to show areas that have larger counts than others.

# Change point shape by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```


# Make all points blue
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```
## A question to try out: 
## What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)? Note, you’ll also need to specify x and y.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  aes(colour = displ < 5)
```

When the aesthetic of 'aes(colour = displ <5) is added as another aesthetic component of the ggplot, any point on the map where the 'displ' (engine size of the car in liters) is smaller than 5 will be a different color. A legend also appears showing that when the engine size is less than 5 the color is recorded as 'TRUE' and if the engine size is larger than 5 it is recorded as 'FALSE.'

**Note** This aesthetic needs to be added on a different line than the 'geom_point' line because it is a different aesthetic command.

# Troubleshooting 

Perhaps the main problem encountered with ggplot is having the + in the wrong place. It has to go at the end of the line (not the start of the next line), which is particularly important if you’re running your script line-by-line.
Example:

```
ggplot(data = mpg) 
+ geom_point(mapping = aes(x = displ, y = hwy)) 
# the + should be on top line
```


# Faceting

To facet your plot using a single variable, use ```facet_wrap()```. Facet wrap syntax is in the function of a formula (kind of like a linear model formula), where the ```~``` dictates which variable you want to subset your data with. 

Note: only use ```facet_wrap()``` for **discrete variables**.


## facet_wrap
Split the plot out by car type (class)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```


## facet_grid
A separate facet for each combination of drive-type (e.g. 4WD) * number of cylinders

If you want to do this with more than one variable, then use ```facet_grid()```. Here you need two variables using ```~``` to split them up.
Facet plot on the combination of two variables using ```facet_grid()```. Facet grid needs two variables separated by a ```~```.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```
Note that there are no occurrences of 5 cylinder 4WDs OR RWD vehicles!



## facet_grid by just row (or column)

Use a ```.``` if you do not want to facet in the rows or column dimension. Note what happens in this case:

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(cyl ~ .)
  #facet_grid(. ~ cyl) # Alternatively


ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  #facet_grid(cyl ~ .)
  facet_grid(. ~ cyl) # Alternatively
```

# Exercise:
Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? 

```{r}
?facet_wrap
```

facet_wrap() wraps a 1d sequence of panels into 2d. This is generally a better use of screen space than facet_grid() because most displays are roughly rectangular.

```nrow``` determines the number of rows while ```ncol``` determines the number of columns. 

**What other options control the layout of the individual panels?**

# Lines
We already used points:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```


However, ggplot2 can use a variety of geom objects to represent the data. Here, we might want to use bar plots, line charts, boxplots and so on. Well we can handle this issue in ggplot directly using a different geom to plot the same data. Here, instead of plotting points, we will use a smooth line. 

# To display the same data as a smooth line fit through the points use geom_smooth().
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))

ggplot(data = mpg) + 
  geom_line(mapping = aes(x = displ, y = hwy))
```
### Question: whey don't we use geom_line() here? What would that look like? ###

We use ```geom_smooth()``` because it creates a line plot that shows the trends of the data set. The smooth function follows a formula for the line it creates (y~x), whereas the ```geom_line()``` function connects a line to each raw data point. This connection of lines is not helpful for this dataset as it only shows the lines of each raw data point instead of the trends of the data set.


So let’s recap. A **geom** is an object that your plot uses to *represent the data*. To change the geom type in your plot, simply change the **geom function** that you add to your plot template. Sometimes you may want to try a few things out, in which case you could use comments to help you remember what worked and what didn’t. 

## Using comments (#)
```{r}
ggplot(data = mpg) + 
  #geom_point(mapping = aes(x = displ, y = hwy)) + # points horrible 
  geom_smooth(mapping = aes(x = displ, y = hwy)) # try smooth line
```

Question: how does R work out where to plot the line??? Can use the chunk output to inform us. Can also use the help menu.

# Changing linetype
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv, color = drv))  # Can also use "lty = " for the same result
```


# Grouping

So far we’ve only made fairly simple plots. But what happens if you want to group objects by a categorical variable (like species or sex or site)? 
Here we will use the ```group``` argument to show that our data is in fact grouped. It won’t add a new legend but can easily be used to vary your plots by the discrete variable.
Set the ```group``` aesthetic to a categorical variable to draw multiple objects.

```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```
 
# Change line colour based on drv value
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE,
  )
```

# Multiple geoms
We already did this one way
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

This is often exactly what we want to do! However, note how the top and second rows are duplicated, meaning if you wanted to change the x variable in your plot, you’d need to change it in several locations!

This is not ideal in a programming sense, and can increase the chance you’ll make an error. Therefore, ggplot allows you to pass these ‘overarching’ mappings to the ```ggplot()``` argument, making them **global mappings** that are applied to *every single subsequent geom*. Much in the same way that the data argument is also global, it is used every time a geom is called.

Here’s an example of where we make the exact same plot as above, but it is programmatically more efficient and makes it easy for you to change a variable that you want to plot on the x or y axis. 


A better way...
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```


# Can still manipulate each geom/layer separately:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

Now we have different colours to indicate the raw data (geom_point) and one colour for the overall relationship (geom_smooth). Neat!


```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```
Question: what does the "se = FALSE" do?

"se = FALSE" turns off the standard error/confidence intervals for the line in the ```geom_smooth``` function.

# Exercise:
1.  What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
2.  Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

```{r}
ggplot(data = mpg, mapping = aes(x= displ, y = hwy)) + 
  #geom_line() #creates line chart
  #geom_boxplot(aes(group = class)) #creates boxplot
  geom_area() #creates area chart

ggplot(data = mpg, mapping = aes(x= displ)) + 
  geom_histogram() #creates histogram

```


3.  Will these two graphs look different? Why/why not?
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

No, the above graphs will not look different because the code is instructing the same map aestetics to be added to the graph. The first code just does it in a more efficient manner.


# Transformations and stats

## Load data
```{r}
data("diamonds")
glimpse(diamonds)
```

# Bar plot
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```


On the x-axis, the chart displays cut, a variable from the ```diamonds``` dataset. On the y-axis, it displays count, but count is not a variable in ```diamonds```! Where does count come from? Many graphs, like scatterplots, plot the **raw values** of your dataset. Other graphs, like bar charts, **calculate new values** to plot:

* bar charts, histograms, and frequency polygons “bin” your data and then plot bin counts (the number of points that fall in each bin).
* smoothers fit a model to your data and then plot predictions from the model.
* boxplots compute a robust summary of the distribution and then display a specially formatted box.

The algorithm used to calculate new values for a graph is called a stat, short for statistical transformation. 

This incredible functionality of ggplot2 will open up a wealth of ways you can visualise your data, without the need to fit statistical models or summarise your dataset!

You can generally use **geoms** and **stats** interchangeably. 
For example, you can recreate the previous plot using ```stat_count()``` instead of ```geom_bar()```.

```{r}
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```

The main reason this is so easy is because *every geom has a default stat* and *every stat has a default geom*, which means you don’t need to worry about what’s going on. 


# Overriding defaults
Make some new data
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo
```


```{r}
ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```
Question: Why did we specify now specify a y axis?

We now specify a y-axis as the frequency because we are changing the default of the graph and we want to make sure the new graph is showing the new dataset we have created with the tribble function.


# What if we want to know the proportion rather than the absolute frequency?
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
```

Question: does anyone get the warning that's in the workbook?
Warning: `stat(prop)` was deprecated in ggplot2 3.4.0.
If so, can use # geom_bar(mapping = aes(x = cut, y = stage(after_stat = prop), group = 1))

**YES**

# Plotting statistical details
```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```


# Aesthetic adjustments adjustments
Another way to boost the way you can convey information with plots using ggplot2 is to use aesthetics like colour or fill to change aspects of bar colours. We already did this once, but there are multiple options available to you, including changing the fill our outline colours.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```
Question: Does anyone notice anything different in the colour scheme of this plot?
(Hint: It's in the viridis colour palette (colour blind friendly), but why is it different from the colour palette we used earlier?)
Check out the difference:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) 

ggplot(data = diamonds, mapping = aes(x = color, y = depth)) + 
  geom_point(mapping = aes(color = cut)) 
```


# Filling by a variable
Now try using these aesthetics to colour by another variable like clarity. Notice how the stacking is done automatically. This is done behind the scenes with a **position** argument.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```


The ability to make **position adjustments** is vital, it allows you to customise your plots in three ways, ```identity``` (raw data), ```fill``` (changes heights) and ```dodge``` (which forces ggplot2 to not put things on top of each other)

* If you use ```position = "identity"```, you will be able to place each object exactly where it falls in the context of the graph. This is vital for point charts like scatter plots but makes a mess in a bar plot situation by showing too much information (a bar plot generally summarises information). So in this case we will need to alter the bar aesthetic.


# To alter transparency (alpha)
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity") #the alpha = 1/5 decreases the transparency so all of the colors representing the clarity can be seen in the barplot
```


# To color the bar outlines with no fill color
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```


# Position adjustments
position = "fill" works like stacking, but makes each set of stacked bars the same height.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```


# position = "dodge" 
Places overlapping objects directly beside one another.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

# Jittering
position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap. This is useful for scatterplots but not barplots.
```{r}
ggplot(data = mpg) + 
  #geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
  geom_point(mapping = aes(x = displ, y = hwy))
```





# The layered grammar of graphics

The sections above have given you the foundation to make almost any type of plot you want to in ggplot2. 
It’s good now to update our template for making a ggplot2. Here it is, with position adjustments, stats, and faceting (recall how facets made us panel plots of subsets?). 
Later in this module we’ll learn about coordinates, which allow for mapping, but for now let’s keep it this simple. Most of the time you won’t need to populate all of these, because ggplot2 has defaults for most of them, but overall you have enough here to build almost any plot.

```
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <FACET_FUNCTION>
```